Comparing vector-based and ACT-R memory models of tag retrieval:\protect\\User-customized hashtag/tag prediction on Twitter and StackOverflow

### 2014-07-15 writeups/dissertation/dissertation.tex
foo

``Stop words'' are commonly used words that co-occur with all tags (i.e., have little to no discriminating power).
It is typically recommended \parencite{Bird2009} to remove stop words such as ``the'', ``a'', ``or'' when building the co-occurrence matrix.
Otherwise the matrix may not scale as well since these low-predictor stop words commonly appear in text.
Also for vector-based systems, memory vectors will become overly saturated with the environment vectors of the commonly-occurring stop words.
\textcite{Sahlgren2008} removed stop words for the random-permutation vector-based model to handle this issue.
\textcite{Jones2007} did as well for the circular convolution model, but also mentioned that an entropy weighting technique has also been used previously to attenuate (rather than remove) stop words.

\textcite{Stanley2013} used the entropy weighting technique recommended in \textcite{Dumais1991} for the StackOverflow dataset to reduce the weight of low-predictor stop words instead of removing them.
This data-driven technique identifies stop words by calculating the discriminating power (i.e., entropy) of each word to predict any tag.
The result maps cleanly to the ACT-R DM theory by modifying each word's attentional weight ($W_{j}$) parameter.
It also seems less arbitrary than explicitly identifying a list of stop words to remove from the analysis.
This technique worked well for the StackOverflow model, and will be further explored for Twitter hashtag prediction.
However, if it becomes computationally infeasible to compute this weighting for Twitter, then the stop words will simply be removed, as was done for previous research.

%FIXME: Add paragraph for freq weighting

@@

How does a user's prior hashtag use influence their likelihood of choosing that hashtag in the future?
The ACT-R theory places strong constraints on the lifetime and growth characteristics of chunks in declarative memory.
The more recently and frequently used chunks have higher prior activations, and are more likely to be needed in the future.
It is an open question if this ACT-R theory of chunk growth and decay accurately reflects hashtag growth and decay for each user on Twitter and StackOverflow.

The full StackOverflow dataset and Twitter popular-users dataset will be used to characterize tag growth and decay for a user on each site.
Afterwards, the predicted results from the ACT-R theory will be compared against the observed data and evaluated.
If the ACT-R theory of decay rate and frequency/recency affects fits the data reasonably well, then the prior activation components for the models developed in this research will be based on this theory.
However, if the theory does not fit hashtag growth and decay for each user, then an alternative model will be proposed that is a better fit, and that model will be used instead.

>>>>

``Stop words'' are commonly used words that co-occur with all tags (i.e., have little to no discriminating power).
It is typically recommended \parencite{Bird2009} to remove stop words such as ``the'', ``a'', ``or'' when building the co-occurrence matrix.
Otherwise the matrix may not scale as well since these low-predictor stop words commonly appear in text.
Also for vector-based systems, memory vectors will become overly saturated with the environment vectors of the commonly-occurring stop words.

Stop words are commonly removed by either filtering based on a pre-determined list or filtering based on the high-frequency words that occur in the dataset.
\textcite{Sahlgren2008} tried both methods to remove stop words for the random-permutation vector-based model, and found that the data-driven approach
(i.e., building the list based on the high-frequency words that occurred in the dataset) worked much better than using a pre-determined list.

\textcite{Jones2007} also filtered the stop words for the circular convolution model,
but also mentioned that an entropy weighting technique \parencite{Dumais1991} has been used previously to attenuate (rather than remove) stop words.
This data-driven technique identifies stop words by calculating the discriminating power (i.e., entropy) of each word to predict any tag.
Each word is then weighted (not removed) based on their discriminating power.

\textcite{Stanley2013} used the entropy weighting technique recommended in \textcite{Dumais1991} for the StackOverflow dataset.
The result maps cleanly to the ACT-R DM theory by modifying each word's attentional weight ($W_{j}$) parameter.
It also seems less arbitrary than explicitly identifying a list of stop words to remove from the analysis.

All three stop word techniques (i.e., pre-determined list, high-frequency filtering, and entropy weighting) were explored for this analysis.

@@

How does a user's prior hashtag use influence their likelihood of choosing that hashtag in the future?
This question was tested by examing how well ACT-R's base-level activation component of declarative memory predicts the tags that authors choose, given each author's past tagging behavior.
The model predicts that the more recently and frequently used chunks have higher prior activations, and are more likely to be needed in the future.

\subsubsection{Base-level ACT-R model}

The base-level activation component from the full ACT-R declarative memory model in Table \ref{tabACTRModel} is included in Table \ref{tabACTRBLLModel}.

\begin{table}[!ht]
  \caption{Base-level component of ACT-R declarative memory}
  \label{tabACTRBLLModel}
  {\tabulinesep=1.2mm
    \begin{tabu}{ll}
      \hline
      Common Name &  Equation \\
      \hline
      Base Level Activation &	$B_{i} = log \sum_{j=1}^{n} {t_{j}}^{-d}$ \\
      Optimized Learning &	$B_{i} = log \frac{n}{1-d} - d * log L$
      \hline
    \end{tabu}
  }
\end{table}

Both terms are methods for computing the base-level activation ($B_{i}$) of a chunk in declarative memory.
The standard form has a time dependence and accounts for the fact that the rate of presentations for a given object can change over time.
The optimized learning form assumes that this rate is constant.

For both equatons, $i$ represents each chunk type (e.g., a word), $j$ is an observation for a chunk (e.g,. the 2nd time that ``a'' appears), and $n$ is the number of times each chunk type has been observed.
Since the optimized learning form assumes a constant rate for each chunk type, it only has to keep track of the time since the first presentation of a chunk type $L$ and the total number of chunks observed for that type. 
It does not have to keep track of the time each specific chunk was observed.
Consequently, the optimized learning form of $B_{i}$ is computationally much less expensive than the standard form, and is actually set to the default in the ACT-R common-lisp implementation.

The decay rate parameter $d$ represents how fast the activation for an individual presentation of a chunk decays from declarative memory.
Higher values mean that chunk activation decays faster, which means that the most recently-presented chunks will have the highest activation, and the total number of times that a chunk type has been presented matters less.
Lower decay rate values flip the importance of frequency and recency: Lower values mean that chunks types that have occured most frequently will have the highest activation,
and activation depends less on the specific chunk types that have been observed in the recent past.

The base-level ($B_{i}$) component in Table \ref{ACTRBLLModel} is different than the $B_{i}$ used in \textcite{Stanley2013} (Table \ref{ModACTRModel}).
\authorcite{Stanley2013} assumed no time dependence, no decay rate, and that the $B_{i}$ is simply a log odds computation of the probability that each chunk type is likely to be presented again, given past frequency.
If a decay rate of 0 is used on the optimized learning form of $B_{i}$, then the equation collapses to $log n_{i}$,
which is a nonlinear transformation of the $B_{i}$ used by \authorcite{Stanley2013} ($log \frac{p_{i}}{1-p_{i}}$).
Both forms maintain the same rank ordering, so they should produce similar results.
So the $B_{i}$ component used by \authorcite{Stanley2013} can be roughly thought of as the standard form of $B_{i}$ when the decay rate is 0 (i.e., a pure frequency model).

\subsubsection{Method}

The full StackOverflow dataset and Twitter popular-users dataset were used to characterize tag growth and decay for a user on each site.
Each user's tag behavior from the popular-users dataset for each site was analyzed.

Multiple samples were taken across a broad range of user types for each dataset to ensure that the model results weren't dependent on specific attributes in a single sample.
For the StackOverflow popular-users dataset, a total of 12 slices of users were taken at different levels of reputation and total number of questions asked.
500 users were sampled that had a reputation higher than \num{500}, \num{1000}, \num{5000}, \num{10000}, \num{50000}, and \num{100000}.
100 users were sampled that had asked a total number of questions higher than \num{50}, \num{100}, \num{200}, \num{300}, \num{400}, \num{500}.

For the Twitter popular-users dataset, 12 slices were also taken, this time at different levels of number of followers and total number of tweets.
100 users were sampled that had a total number of followers higher than \num{1000}, \num{5000}, \num{10000}, \num{100000}, \num{1000000}, and \num{10000000},
and a total number of tweets higher than \num{100}, \num{500}, \num{1000}, \num{5000}, \num{10000}, and \num{50000}.

\subsubsection{Results}
